Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2015-10-19 patchy-selection-paper-for-PLoS.tex
final?

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
surrounded by ``branches'',
% and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Fig~\ref{fig:branching_decomp}.
We can understand this with a decomposition described by \citet{geiger1999elementary}.
% (see also \citet{chauvin1991growing}). 
% (lyons1995conceptual deals with conditioned on eventual nonextinction)
When we condition on survival until $t$, we condition on survival of at least one trunk lineage
(the long red lineage in Fig \ref{fig:branching_decomp}).
Given this lineage, the other reproduction events are undistinguished --
% each individual along the trunk lineage must give birth to at least one offspring that survives until time $t$,
% and all other reproduction events are unconditioned.
% In other words, 
the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent, unconditioned branches that may or may not survive until $t$.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then
\begin{align} \label{eqn:Zt_decomposition}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is independent with distribution 
%% equation (2.3) in Geiger:
%% $\P\{ W_s = k\} = \sum_{j=k}^\infty \P\{ W = j \} k_e(t-s-1)^{j-k} (1-k_e(t-s-1))/(1-k_e(t-s))$, 
%% and equation (2.5) in Geiger:
that tends to $\P\{W_s=k\} \simeq \P\{W\ge k\}/\E[W]$ as $s \to \infty$,
where $W$ is the unconditioned offspring number distribution.  \linelabel{ll:branching_process_fix}
In particular, $\E[W_s-1] \simeq \E[W(W-1)]/\E[W]$;
if $W$ is Poisson then this is equal to $e^{-s_m}$.
Since each subfamily $Z^{(s,k)}$ is a branching process with mean offspring number less than one,
most of these die out quickly,
and so $Z_t$ is composed of a cluster of recently split subfamilies.
If we take the expectation of \eqref{eqn:Zt_decomposition},
since $\E[Z_t] = e^{-s_m t}$,
we get the limiting mean family size conditioned on nonextinction:
\begin{align} \label{eqn:mean_K}
    \E[K] 
    = 1 + \sum_{s=0}^\infty \E[W_s-1] e^{-s_m t}
    \approx \E[W(W-1)] / ( \E[W] s_m  )
\end{align}

@@

The remaining term is $q(S)$, the integral of equation \eqref{eqn:eqfreq} over the patch.
Since the patches are at distance $R$,
if the new patch has width smaller than $\sigma/\sqrt{2s_m}$, 
then $q(S)$ is approximately $q(R)$ multiplied by the area of the new patch.
If $S$ is larger, then we should use the area of the closest strip
of width $\sigma/\sqrt{2s_m}$ to the original patch;
see Fig \ref{fig:branching_decomp}, and section \secref{apx:qS} for more details.
In either case, denote this area $A'$;
for numerical examples we compute $q(S)$ directly.

@@

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      % \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{Observed versus predicted time until adaptation.}
      \textbf{(left)} 
      Time until adaptation by mutation in an isolated patch for different combinations
      of selection coefficient on patches ($s_m$), 
      population density ($\rho$),
      mutation rate ($\mu$),
      and distance between patches ($R$),
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (horizontal axis).
      In all simulations, $s_p=0.01$.
      \textbf{(right)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate}, with $C=5$ (horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:mutation_params} and \ref{stab:migration_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
      Only parameter combinations to which the theory here is applicable were used
      (precisely, if
      $R > \sigma/\sqrt{s_m}$, 
      patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_p})$, 
      and $s_m > 2 \sigma \rho$);
      the remaining combinations are shown in 
      S\ref{sfig:sim_migration_times} and S\ref{sfig:sim_mutation_times} Fig.
  }   \label{fig:sim_times}
\end{figure}

@@

We tested this using the same simulations as Fig \ref{fig:sim_times},
by using the empirical distributions of the respective times to adaptation
to estimate the probability,
for each parameter combination,
that a new, successful mutation appears 
before a successful migrant arrives from another, already adapted patch.
The results are compared to \eqref{eqn:parallel_prob}
in Fig~\ref{fig:sim_probs}.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      % \includegraphics{prob-mutation-compared}
  \end{center}
  \caption{
      \textbf{Probability that two patches 
      evolve by independent mutations (i.e., convergently),
      as a function of their geographic separation ($R$)
      and the strength of selection between patches ($s_m$).}
      The background shades of grey and contours
      give the probability predicted by theory (equation~\eqref{eqn:parallel_prob}, with $C=5$).
      Points correspond to parameter combinations of the simulations in Fig \ref{fig:sim_times},
      with the shade of grey and the red numeric label indicating the same probability
      estimated from these simulations (see text for details).
      The two densities, $\rho=100$ and $\rho=1000$ would lie directly on top of each other, so the $\rho=1000$ points have been shifted vertically.
  }   \label{fig:sim_probs}
\end{figure}

@@

Here we make a more precise argument to back up expression \eqref{eqn:migrate} for the migration rate.
The argument made above in section \secref{ss:heuristics} applies to general migration mechanisms,
since it relies only on a decomposition of the migrant families upon hitting the new patch;
but it is also imprecise in subtle ways
that are difficult to formalize.
Here we take a somewhat different tack,
supposing that it suffices to model the spatial movement of a migrant family
by following only the motion of the ``trunk''
(i.e., the red line in Fig \ref{fig:branching_decomp}),
and supposing this motion is Brownian, with variance $\sigma^2$ per generation.
(Recall $\sigma$ is the dispersal distance.)
We then use facts about Brownian motion and branching processes, 
to compute more precise versions of equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate}.
\linelabel{rr:moved_argument}

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are
given by a spatial branching process (as depicted in Fig \ref{fig:branching_decomp}).
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\partial B(r_0)$ its boundary.
Denote by $\gamma(x)$ the mean rate of outflux of migrant families from a point $x \in \partial B(r_0)$,
i.e., the time-averaged density of individuals near a point $x$ in $\partial B(r_0)$ 
that are the founders of new migrant families.
Expressions \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate} are a simple product of the ``outflux of families'',
when in fact they should be an integral of $\gamma(x)$ over possible locations. \linelabel{rr:awkward}
However,
it will turn out that the integrand of equation \eqref{eqn:gestalt_migrate} 
is well-approximated by a constant multiple of the integrand of \eqref{eqn:gestalt_q}.

@@

% \ifsubmissionversion
% \clearpage
% \renewcommand*\listfigurename{Figure Legends}
% \listoffigures
% \fi
% 
% \clearpage
% 
% % endfloat uses different names
% \renewcommand{\thesfigure}{S\arabic{sfigure}}
% \renewcommand{\thetable}{S\arabic{table}}
% \ifsubmissionversion
%   \renewcommand{\thepostsfigure}{S\arabic{postsfigure}}
%   % \renewcommand{\thefigure}{S\arabic{figure}}
%   % \renewcommand{\thepostfigure}{S\arabic{postfig}}
%   % \setcounter{postfig}{0}
%   \renewcommand{\theposttable}{S\arabic{posttable}}
%   \setcounter{posttable}{0}
% \fi
% 
% %% list of supplementary figures
% \ifsubmissionversion
%   \renewcommand*\listsfigurename{Supplementary Figure Legends}
%   \listofsfigures
%   \renewcommand*\listtablename{Supplementary Table Legends}
%   \listoftables
%   \pagebreak
% \fi
% 
% \appendix
% \section*{Supporting Information}
% 
% \setcounter{secnumdepth}{2}

@@

\subsection*{S1 Fig}
\label{sfig:sim_migration_times}

@@

\subsection*{S2 Fig}
\label{sfig:sim_mutation_times}

@@

% example trace plots

\subsection*{S3 Fig}
\label{sfig:sims_1}

@@

\subsection*{S4 Fig}
\label{sfig:sims_2}

@@

\subsection*{S5 Fig}
\label{sfig:sims_3}

@@

\subsection*{S6 Fig}
\label{sfig:sims_4}

@@

\subsection*{S7 Fig}
\label{sfig:sims_5}

@@

\subsection*{S8 Fig}
\label{sfig:sims_6}

% % sm=0.01, R=80
% \begin{sfigure}
%   \begin{center}
%     % \includegraphics{example-migration-sims/42080-r1-501-sb0_01-sm-0_01-N4000-pophistory-run}
%     % \includegraphics{example-migration-sims/82698-r1-501-sb0_01-sm-0_01-N1000-pophistory-run}
%     % \includegraphics{example-migration-sims/97545-r1-501-sb0_01-sm-0_01-N100-pophistory-run}
%   \end{center}
%   \caption{
\textbf{Randomly chosen simulations of adaptation by migration
with $s_m=0.01$, $R=80$, $\sigma\approx 1$, and $\rho$ varying.}
    On the left of each is a space-time heatmap of the local frequency of $B$ alleles;
    and on the right are twenty-five curves showing the frequencies of $B$ at evenly spaced time points
    (i.e., each line represents a vertical slice through the plot on the left);
    dotted black lines indicate the patches where $B$ is advantageous.
%   } \label{sfig:sims_6}
% \end{sfigure}

\subsection*{S9 Fig}
\label{sfig:sims_7}

% % sm=0.001, R=80
% \begin{sfigure}
%   \begin{center}
%     % \includegraphics{example-migration-sims/88079-r1-501-sb0_01-sm-0_001-N4000-pophistory-run}
%     % \includegraphics{example-migration-sims/3464-r1-501-sb0_01-sm-0_001-N1000-pophistory-run}
%     % \includegraphics{example-migration-sims/37774-r1-501-sb0_01-sm-0_001-N100-pophistory-run}
%   \end{center}
%   \caption{
\textbf{Randomly chosen simulations of adaptation by migration
with $s_m=0.001$, $R=80$, $\sigma\approx 1$, and $\rho$ varying.}
    On the left of each is a space-time heatmap of the local frequency of $B$ alleles;
    and on the right are twenty-five curves showing the frequencies of $B$ at evenly spaced time points
    (i.e., each line represents a vertical slice through the plot on the left);
    dotted black lines indicate the patches where $B$ is advantageous.
%   } \label{sfig:sims_7}
% \end{sfigure}

\subsection*{S10 Fig}
\label{sfig:sims_8}

>>>>

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
surrounded by ``branches'',
% and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Fig~\ref{fig:branching_decomp}.
We can understand this with a decomposition described by \citet{geiger1999elementary}.
% (see also \citet{chauvin1991growing}). 
% (lyons1995conceptual deals with conditioned on eventual nonextinction)
When we condition on survival until $t$, we condition on survival of at least one trunk lineage
(the long red lineage in Fig~\ref{fig:branching_decomp}).
Given this lineage, the other reproduction events are undistinguished --
% each individual along the trunk lineage must give birth to at least one offspring that survives until time $t$,
% and all other reproduction events are unconditioned.
% In other words, 
the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent, unconditioned branches that may or may not survive until $t$.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then
\begin{align} \label{eqn:Zt_decomposition}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is independent with distribution 
%% equation (2.3) in Geiger:
%% $\P\{ W_s = k\} = \sum_{j=k}^\infty \P\{ W = j \} k_e(t-s-1)^{j-k} (1-k_e(t-s-1))/(1-k_e(t-s))$, 
%% and equation (2.5) in Geiger:
that tends to $\P\{W_s=k\} \simeq \P\{W\ge k\}/\E[W]$ as $s \to \infty$,
where $W$ is the unconditioned offspring number distribution.  \linelabel{ll:branching_process_fix}
In particular, $\E[W_s-1] \simeq \E[W(W-1)]/\E[W]$;
if $W$ is Poisson then this is equal to $e^{-s_m}$.
Since each subfamily $Z^{(s,k)}$ is a branching process with mean offspring number less than one,
most of these die out quickly,
and so $Z_t$ is composed of a cluster of recently split subfamilies.
If we take the expectation of \eqref{eqn:Zt_decomposition},
since $\E[Z_t] = e^{-s_m t}$,
we get the limiting mean family size conditioned on nonextinction:
\begin{align} \label{eqn:mean_K}
    \E[K] 
    = 1 + \sum_{s=0}^\infty \E[W_s-1] e^{-s_m t}
    \approx \E[W(W-1)] / ( \E[W] s_m  )
\end{align}

@@

The remaining term is $q(S)$, the integral of equation \eqref{eqn:eqfreq} over the patch.
Since the patches are at distance $R$,
if the new patch has width smaller than $\sigma/\sqrt{2s_m}$, 
then $q(S)$ is approximately $q(R)$ multiplied by the area of the new patch.
If $S$ is larger, then we should use the area of the closest strip
of width $\sigma/\sqrt{2s_m}$ to the original patch;
see Fig~\ref{fig:branching_decomp}, and section \secref{apx:qS} for more details.
In either case, denote this area $A'$;
for numerical examples we compute $q(S)$ directly.

@@

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      % \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{Observed versus predicted time until adaptation.}
      \textbf{(left)} 
      Time until adaptation by mutation in an isolated patch for different combinations
      of selection coefficient on patches ($s_m$), 
      population density ($\rho$),
      mutation rate ($\mu$),
      and distance between patches ($R$),
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (horizontal axis).
      In all simulations, $s_p=0.01$.
      \textbf{(right)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate}, with $C=5$ (horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:mutation_params} and \ref{stab:migration_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
      Only parameter combinations to which the theory here is applicable were used
      (precisely, if
      $R > \sigma/\sqrt{s_m}$, 
      patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_p})$, 
      and $s_m > 2 \sigma \rho$);
      the remaining combinations are shown in 
      \ref{sfig:sim_migration_times} and \ref{sfig:sim_mutation_times} Fig.
  }   \label{fig:sim_times}
\end{figure}

@@

We tested this using the same simulations as Fig~\ref{fig:sim_times},
by using the empirical distributions of the respective times to adaptation
to estimate the probability,
for each parameter combination,
that a new, successful mutation appears 
before a successful migrant arrives from another, already adapted patch.
The results are compared to \eqref{eqn:parallel_prob}
in Fig~\ref{fig:sim_probs}.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      % \includegraphics{prob-mutation-compared}
  \end{center}
  \caption{
      \textbf{Probability that two patches 
      evolve by independent mutations (i.e., convergently),
      as a function of their geographic separation ($R$)
      and the strength of selection between patches ($s_m$).}
      The background shades of grey and contours
      give the probability predicted by theory (equation~\eqref{eqn:parallel_prob}, with $C=5$).
      Points correspond to parameter combinations of the simulations in Fig~\ref{fig:sim_times},
      with the shade of grey and the red numeric label indicating the same probability
      estimated from these simulations (see text for details).
      The two densities, $\rho=100$ and $\rho=1000$ would lie directly on top of each other, so the $\rho=1000$ points have been shifted vertically.
  }   \label{fig:sim_probs}
\end{figure}

@@

Here we make a more precise argument to back up expression \eqref{eqn:migrate} for the migration rate.
The argument made above in section \secref{ss:heuristics} applies to general migration mechanisms,
since it relies only on a decomposition of the migrant families upon hitting the new patch;
but it is also imprecise in subtle ways
that are difficult to formalize.
Here we take a somewhat different tack,
supposing that it suffices to model the spatial movement of a migrant family
by following only the motion of the ``trunk''
(i.e., the red line in Fig~\ref{fig:branching_decomp}),
and supposing this motion is Brownian, with variance $\sigma^2$ per generation.
(Recall $\sigma$ is the dispersal distance.)
We then use facts about Brownian motion and branching processes, 
to compute more precise versions of equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate}.
\linelabel{rr:moved_argument}

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are
given by a spatial branching process (as depicted in Fig~\ref{fig:branching_decomp}).
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\partial B(r_0)$ its boundary.
Denote by $\gamma(x)$ the mean rate of outflux of migrant families from a point $x \in \partial B(r_0)$,
i.e., the time-averaged density of individuals near a point $x$ in $\partial B(r_0)$ 
that are the founders of new migrant families.
Expressions \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate} are a simple product of the ``outflux of families'',
when in fact they should be an integral of $\gamma(x)$ over possible locations. \linelabel{rr:awkward}
However,
it will turn out that the integrand of equation \eqref{eqn:gestalt_migrate} 
is well-approximated by a constant multiple of the integrand of \eqref{eqn:gestalt_q}.

@@

% \ifsubmissionversion
% \clearpage
% \renewcommand*\listfigurename{Figure Legends}
% \listoffigures
% \fi
% 
% \clearpage
% 
% % endfloat uses different names
% \renewcommand{\thesfigure}{S\arabic{sfigure}}
% \renewcommand{\thetable}{S\arabic{table}}
% \ifsubmissionversion
%   \renewcommand{\thepostsfigure}{S\arabic{postsfigure}}
%   % \renewcommand{\thefigure}{S\arabic{figure}}
%   % \renewcommand{\thepostfigure}{S\arabic{postfig}}
%   % \setcounter{postfig}{0}
%   \renewcommand{\theposttable}{S\arabic{posttable}}
%   \setcounter{posttable}{0}
% \fi
% 
% %% list of supplementary figures
% \ifsubmissionversion
%   \renewcommand*\listsfigurename{Supplementary Figure Legends}
%   \listofsfigures
%   \renewcommand*\listtablename{Supplementary Table Legends}
%   \listoftables
%   \pagebreak
% \fi
%

\appendix
\renewcommand{\thesubsection}{S\arabic{subsection}}
\setcounter{secnumdepth}{2}

@@

% S1
\subsection{Fig}
\label{sfig:sim_migration_times}

@@

% S2 
\subsection{Fig}
\label{sfig:sim_mutation_times}

@@

% example trace plots
% S3 
\subsection{Fig}
\label{sfig:sims_1}

@@

% S4 
\subsection{Fig}
\label{sfig:sims_2}

@@

% S5 
\subsection{Fig}
\label{sfig:sims_3}

@@

% S6 
\subsection{Fig}
\label{sfig:sims_4}

@@

% S7 
\subsection{Fig}
\label{sfig:sims_5}

@@

% S8 
\subsection{Fig}
\label{sfig:sims_6}

% % sm=0.01, R=80
% \begin{sfigure}
%   \begin{center}
%     % \includegraphics{example-migration-sims/42080-r1-501-sb0_01-sm-0_01-N4000-pophistory-run}
%     % \includegraphics{example-migration-sims/82698-r1-501-sb0_01-sm-0_01-N1000-pophistory-run}
%     % \includegraphics{example-migration-sims/97545-r1-501-sb0_01-sm-0_01-N100-pophistory-run}
%   \end{center}
%   \caption{
\textbf{Randomly chosen simulations of adaptation by migration
with $s_m=0.01$, $R=80$, $\sigma\approx 1$, and $\rho$ varying.}
    On the left of each is a space-time heatmap of the local frequency of $B$ alleles;
    and on the right are twenty-five curves showing the frequencies of $B$ at evenly spaced time points
    (i.e., each line represents a vertical slice through the plot on the left);
    dotted black lines indicate the patches where $B$ is advantageous.
%   } \label{sfig:sims_6}
% \end{sfigure}
% S9 
\subsection{Fig}
\label{sfig:sims_7}

% % sm=0.001, R=80
% \begin{sfigure}
%   \begin{center}
%     % \includegraphics{example-migration-sims/88079-r1-501-sb0_01-sm-0_001-N4000-pophistory-run}
%     % \includegraphics{example-migration-sims/3464-r1-501-sb0_01-sm-0_001-N1000-pophistory-run}
%     % \includegraphics{example-migration-sims/37774-r1-501-sb0_01-sm-0_001-N100-pophistory-run}
%   \end{center}
%   \caption{
\textbf{Randomly chosen simulations of adaptation by migration
with $s_m=0.001$, $R=80$, $\sigma\approx 1$, and $\rho$ varying.}
    On the left of each is a space-time heatmap of the local frequency of $B$ alleles;
    and on the right are twenty-five curves showing the frequencies of $B$ at evenly spaced time points
    (i.e., each line represents a vertical slice through the plot on the left);
    dotted black lines indicate the patches where $B$ is advantageous.
%   } \label{sfig:sims_7}
% \end{sfigure}
% S10 
\subsection{Fig}
\label{sfig:sims_8}
